### 2017-12-12 submission/manuscript.tex
final edits of final read through after pat rd 2 suggestions

\textbf{Background.} An increasing body of literature suggests that
there is a crucial role for the microbiota in colorectal cancer (CRC)
pathogenesis. Important drivers within this context have ranged from
individual microbes to the whole community. Our study expands on a
recent meta-analysis investigating microbial biomarkers for CRC by
testing the hypothesis that the bacterial community is an important
driver of both early (adenoma) and late (carcinoma) stage of disease. To
test this hypothesis we examined both feces (n = 1737) and tissue (492
total samples from 350 individuals) across 14 different studies.

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness resulted in small
increases in relative risk for adenoma (P-value = 0.032) and carcinoma
(P-value = 0.00034) while the reduction in Shannon diversity only
resulted in an increased relative risk for carcinoma (P-value = 0.0047).
Previously associated colorectal cancer genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
followed a similar pattern, with their presence significantly increasing
the relative risk for carcinoma (P-value \textless{} 0.05) but not
adenoma (P-value \textgreater{} 0.05) with the exception of
\emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
only CRC associated genera to build a prediction model resulted in
higher classification success based on Area Under the Curve (AUC) for
both adenoma and carcinoma using fecal and tissue samples. The most
important OTUs for these models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Overall, there were less associations between the microbiota
and adenoma versus carcinoma and one reason why this may be is that most
studies were only adequatly powered for large effect sizes.

\textbf{Conclusions.} This data provides support for the importance of
the bacteral community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
CRC pathogenesis shows stronger associations between carcinoma then
adenoma. A strong reason for this may be in part due to the low power to
detect more subtle changes in the majority of studies that have been
performed to date.

@@

Colorectal cancer (CRC) is a growing world-wide health problem {[}1{]}
in which the microbiota has been purported to play an active role in
disease pathogenesis {[}2{]}. Numerous studies have shown the importance
of both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case/control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis, along
with other studies, focused on identifying biomarkers or individual
microbes but do not critically investigate how the community changes in
CRC.

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}12,13{]}. One advantage of tissue is that stool
possibly represents an agregate community of the overall
gastrointestinal (GI) tract and sampling the adenoma or carinoma
directly could more accurately identify important microbes for CRC.
Sampling the tissue directly may provide clearer answers but is not
without limitations. Due to bowel prep the communities left for sampling
may not be reflective of the resident microbiota, but rather a
collection of what is able to keep adhered to the mucosa. Additionally,
these samples contain more host DNA, potentially limiting the types of
analysis that can be done. It is well known that low biomass samples can
be very difficult to work with and results can end up being study
dependent due to the randomness of contamination {[}14{]}. Due to these
many differences that could arise between stool and tissue, one question
our meta-analysis aims to answer is whether there are consistent
patterns that emerge across studies regardless of whether they used
stool or tissue samples.

This intense focus on identifying biomarkers has found strong
associations with resident mouth microbes as potential CRC-associated
microbes in the GI tract {[}15--17{]}. The main bacteria of interest,
arising from this set of microbes, has been those within the
\emph{Fusobacterium} genus. Yet, the question remains as to whether or
not this is indeed the most important genera to be focusing on, since
many microbiota-based studies typically have identified a collection of
oral microbes rather than single species from a single genera
{[}16,17{]}. Based on this discrepency, the second question we want to
answer with this meta-analysis is if there is one dominant
CRC-associated genera that can be identified across studies.

@@

In comparison to the previous meta-analysis, this study significantly
increases the total stool samples investigated, examines differences
between stool and tissue microbiota in the context of CRC, and takes a
more community centric approach rather than a biomarker focused approach
to investigating commonalities across study for the microbiota and CRC
severity. Importantly, this community centric approach could provide
valuable insights into the importance of accounting for the community in
CRC disease not previously provided by earlier meta-analysis studies
{[}11{]}.

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether diversity changes throughout
disease (control to adenoma to carcinoma) and if it results in an
increased relative risk (RR) for adenoma or carcinoma. We then assessed
how common CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the relative risk of adenoma or carcinoma. Next, using Random
Forest models, we analyzed whether the full community or only the
CRC-associated genera resulted in better model classification based on
the area under the curve (AUC). Our results suggest that the community
changes as disease severity worsens and that this community is important
for disease classification. However, since the changes in community were
subtle for adenoma we also examined what effect and sample size the
studies that were used were adequately powered for. Although we analyzed
data sets that sampled large numbers of individuals, our results
indicate the individual studies were underpowered for detecting effect
size differences of 10\% or below between the case and control groups.

@@

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} By power transforming and Z-score normalizing the
\(\alpha\)-diversity metrics for the entire data set we assessed whether
there were any broad scale community differences that could be detected
as disease severity worsened. Using linear mixed-effect models to
control for study, resampling of the same individual (tissue studies
only), and variable region, there was a significant decrease from
control to adenoma to carcinoma for both evenness and Shannon diversity
in stool (P-value = 0.025 and 0.043, respectively) but not for tissue
(P-value \textgreater{} 0.05) {[}Figure 1{]}. We next tested whether
these detectable differences in community resulted in significant
increases in RR. For fecal samples, a decrease in evenness resulted in a
significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
P-value = 0.00034 and adenoma (RR = 1.16 (1.01 - 1.34), P-value = 0.032)
while a decrease in Shannon diversity only increased the RR for
carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
Interestingly, for both adenoma and carcinoma there was no increase in
RR within tissue samples for any alpha diversity metric investigated
{[}Table S1-S3{]}.

Using the Bray-Curtis distance metric, there was a significant
difference across studies in the bacterial community of fecal samples
between carcinoma and controls, but not adenoma and controls {[}Table S4
\& S5{]}. For studies with unmatched tissue samples a similar trend was
observed {[}Table S3 \& S4{]} while studies with tissue samples from the
same individual (matched) had no differences {[}Table S6 \& S7{]}.

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
metrics, we assessed if the presence of specific genera resulted in a
higher RR for both stool and tissue. To investigate this we analyzed the
classically associated CRC genera, \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}
for an increase in RR. The majority of CRC-associated genera for both
feces and tissue had a significantly increased RR for carcinoma but not
for adenoma {[}Figure 3{]}. The RR effect size was greater for stool (RR
range = 1.62 - 2.37) than for tissue (RR range = 1.21 - 1.81). This
decrease may be explained by the fact that the tissue analysis included
matched samples from the same individual. In fecal samples, the RR for
carcinoma due to the prescence of CRC-associated genera was greater than
either the RR associated with evenness or Shannon diversity {[}Figure 2
\& 3{]}. Additionally, the RR of carcinoma continuously increased as
individuals tested positive for more CRC-associated genera {[}Figure 3B
\& 3D{]}.

There were two significant measures for increased RR of adenoma when
investigating CRC-associated genera in stool: 1) Having a higher then
median value of \emph{Porphyromonas} (P-value = 0.023) and 2) whether
samples were positive for three CRC associated genera (P-value = 0.022)
{[}Figure 3A{]}. With tissue, there were three significant measures for
an increased RR of adenoma: 1) being positive for one CRC-associated
genera (P-value = 0.032), 2) being positive for two CRC associated
genera (P-value = 0.008), and 3) being positive for four CRC associated
genera (P-value = 0.039) {[}Figure 3C{]}.

@@

Our first approach using genus based models showed an AUC decrease in
model classification on the training set for both stool and tissue
studies {[}Figure S2-S3{]}. With respect to the test sets, comprised of
genera data from other studies, both the all genera model and
CRC-associated models had a similar ability to detect adenomas or
carcinomas {[}Figure S4-S6{]}. Two interesting but seperate general
observations from these models were that: 1) classification of adenomas
was lower than carcinomas for both tissue and stool and 2) AUC for the
classification of carcinoma was consistently lower for the tissue models
than the stool models {[}Figure S4-S6{]}.

@@

In stool the most common genera in the genus based models belonged
mostly to resident genera such as \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} {[}Figure 6A \& B{]}. With
respect to the CRC-associated genera, \emph{Fusobacterium} was the only
genus present in adenoma while all four were present in carcinoma
{[}Figure 6A \& B{]}. Conversely, none of these CRC-associated genera
were present in the majority of studies. When we move onto the OTU based
models, the adenoma OTU models had OTUs that classified as
\emph{Ruminococcaceae} or \emph{Roseburia} present in the top 10 OTUs
for the vast majority of studies {[}Figure 6C{]}. \emph{Ruminococcaceae}
was also present in the top 10 in some studies for the carcinoma OTU
models, but it was \emph{Bacteroides} that was present in the
overwhelming majority of the carcinoma OTU stool models {[}Figure 6D{]}.

@@

Our study identifies clear differences in diversity both at the
community level and within individual genera that are present in
individuals with CRC versus those without the disease {[}Figure 1-3{]}.
Although there was a step-wise decrease in diversity from control to
adenoma to carcinoma, this did not translate into large effect sizes for
the relative risk of either of these two. Even though CRC-associated
genera increase the RR of carcinoma, they do not consistently increase
the relative risk of adenoma. This information suggests that these
specific genera are strongly associated with carcinoma but may not be
associated with adenoma. Although CRC-associated genera are asscoiated
with carcinomas, our data show that by using the whole community,
classification models perform better than when only the CRC-associated
genera are included. CRC-associated genera are clearly important to
carcinoma pathogenesis but accounting for the community in which these
microbes exist can drastically increase the ability of models to make
predictions. These observations suggest that small localized changes
within the community on tissue may be occuring, that are important in
early disease progression of CRC, and that this process may not directly
involve CRC-associated genera.

The data presented herein supports the driver-passenger model of the
microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
carcinoma but not necessarily adenoma. The central idea of the model is
that a single bacterium initiates an environment in which other
non-resident microbes may then be able to colonize, creating a vicious
cycle that is conducive for CRC. Both the drastically increased RR of
CRC-associated genera versus \(\alpha\)-diversity metrics for carcinoma
and increasing RR with more CRC-associated genera positivity are highly
supportive of this model. However, the initial establishment of the
driver within the system appears to be dependent on the current
community. This is supported by our finding that when adding the
community context to our models in addition to the CRC-associated
genera, the model AUC increases {[}Figure 4 \& 5{]}. Conversely, using
the present data, it is less likely that adenoma development fits this
model. The changes that occur at this timepoint are small and possibly
focal to the adenoma itself. The stepwise decrease in diversity suggests
that the adenoma community is not normal but has changed subtly
{[}Figure 1{]}. Although there appears to be localized changes that do
depend on the driver-passenger model, as supported by an increased RR
for one, two, and four positive CRC-associated genera in tissue
{[}Figure 3C{]}, there may be other processes at play that ultimately
exacerbate the condition from a subtle localized change, to a change in
the global community. The poor performance of the Random Forest models
for classifying adenoma based only on the microbiota would suggest that
this is the case. One potential hypothesis from these observations is
that at early stages of the disease, how the host interacts with these
subtle changes is what ultimately leads to a thoroughly dysfunctional
community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be disciminative across the different studies for
both adenoma and carcinoma models {[}Figure 6{]}. This data would
suggest that whether the non-resident bacterium is \emph{Fusobacteria}
or \emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. These observations would
also suggest that initial changes in the resident community,
specifically to \emph{Bacteroides}, \emph{Ruminococcaceae},
\emph{Ruminococcus}, and \emph{Roseburia}, carry on from adenoma to
carcinoma. Based on these observations, it is possible to hypothesize
that the early changes in the community may give rise to initial polyp
formation via interactions with the host and not necessarily via
interactions with CRC-assciated genera. These changes then create new
niches in which any one of the CRC-associated genera could gain a
foothold, exacerbating the initial changes in community and facilitating
the transition from adenoma to carcinoma via a driver-passenger type
mechanism.

@@

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, \emph{et
al.} and Vogtmann, \emph{et al.} {[}29,30{]}. Additional studies not
mentioned in the reviews were obtained based on the authors' knowledge
of the literature. Studies that used tissue or feces as their sample
source for 454 or Illumina 16S rRNA gene sequencing analysis were
included. Only data sets that had sequences available for analysis were
included. Some studies did not have publically available sequences or
did not have metadata in which the authors were able to share and were
excluded. After these filtering steps, the following studies remained:
Ahn, \emph{et al.} {[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim,
\emph{et al.} {[}22{]}, Burns, \emph{et al.} {[}27{]}, Chen, \emph{et
al.} {[}19{]}, Dejea, \emph{et al.} {[}24{]}, Flemer, \emph{et al.}
{[}12{]}, Geng, \emph{et al.} {[}28{]}, Hale, \emph{et al.} {[}18{]},
Kostic, \emph{et al.} {[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy,
\emph{et al.} {[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et
al.} {[}23{]}, and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]}
study was not included becasue the 90 individuals analyzed within the
study are contained within the larger Baxter study {[}16{]}.
Additionally, after sequence processing all the case samples for the
Kostic study only had 100 or less sequences remaining and was not used.
This left a total of 14 studies for which analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}26{]} and for four studies was obtained directly from
the authors {[}12,18,23,25{]}. Each study was processed using the mothur
(v1.39.3) software program {[}32{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identifed and
removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}34{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.2) software package {[}35{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
normalized using the car (v2.1.5) package {[}37{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}38{]} to correct for study,
repeat sampling of individuals (tissue only), and variable region
effects. Relative risk was analyzed using both the epiR (v0.9.87) and
metafor (v2.0.0) packages {[}39,40{]} by assessing how many with and
without disease were above and below the overall median value within the
specific study. Relative risk significance testing utilized the
chi-squred test. \(\beta\)-diversity differences utilized a Bray-Curtis
distance matrix and PERMANOVA executed with the vegan (v2.4.4) package
{[}41{]}. Random Forest models were built using both the caret (v6.0.77)
and randomForest (v4.6.12) packages {[}42,43{]}. Power analysis and
estimations were made using the pwr (v1.2.1) and statmod (v1.4.30)
packages {[}44,45{]}. All figures were created using both ggplot2
(v2.2.1) and gridExtra (v2.3) packages {[}46,47{]}.

>>>>

\textbf{Background.} An increasing body of literature suggests that
there is a crucial role for the microbiota in colorectal cancer (CRC)
pathogenesis. Important drivers within this context have ranged from
individual microbes to the whole community. Our study expands on a
recent meta-analysis investigating microbial biomarkers for CRC by
testing the hypothesis that the bacterial community has important
associations to both early (adenoma) and late (carcinoma) stage disease.
To test this hypothesis we examined both feces (n = 1737) and tissue
(492 total samples from 350 individuals) across 14 different studies.

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness resulted in small
increases in relative risk for adenoma (P-value = 0.032) and carcinoma
(P-value = 0.00034) while the reduction in Shannon diversity only
resulted in an increased relative risk for carcinoma (P-value = 0.0047).
Previously associated colorectal cancer genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
followed a similar pattern, with their presence significantly increasing
the relative risk for carcinoma (P-value \textless{} 0.05) but not
adenoma (P-value \textgreater{} 0.05) with the exception of
\emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
only CRC-associated genera to build a prediction model resulted in
higher classification success, based on Area Under the Curve (AUC), for
both adenoma and carcinoma using fecal and tissue samples. The most
important OTUs for these models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Overall, there were less associations between the microbiota
and adenoma and one reason why this may be is that most studies were
only adequately powered for large effect sizes.

\textbf{Conclusions.} This data provides support for the importance of
the bacterial community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
CRC pathogenesis shows stronger associations between carcinoma then
adenoma. One reason for this may be in part due to the low power to
detect more subtle changes in the majority of studies that have been
performed to date.

@@

Colorectal cancer (CRC) is a growing world-wide health problem {[}1{]}
in which the microbiota has been purported to play an active role in
disease pathogenesis {[}2{]}. Numerous studies have shown the importance
of both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case/control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis focused
on identifying biomarkers or individual microbes but did not critically
investigate how the community changes in CRC.

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}12,13{]}. Sampling the tissue directly may
provide clearer answers but is not without limitations. Due to bowel
prep the communities left for sampling may not be reflective of the
resident microbiota, but rather a collection of what is able to keep
adhered to the mucosa. Additionally, these samples contain more host
DNA, potentially limiting the types of analysis that can be done. It is
well known that low biomass samples can be very difficult to work with
and results can end up being study dependent due to the randomness of
contamination {[}14{]}. Due to these many differences that could arise
between stool and tissue, one question our meta-analysis aims to answer
is whether there are consistent patterns that emerge across studies
regardless of whether they used stool or tissue samples.

This intense focus on identifying biomarkers has found strong
associations with resident mouth microbes as potential CRC-associated
microbes in the gastrointestinal (GI) tract {[}15--17{]}. The main
bacteria of interest, arising from this set of microbes, has been those
within the \emph{Fusobacterium} genus. Yet, the question remains as to
whether or not this is indeed the most important genera to be focusing
on, since many microbiota-based studies typically have identified a
collection of oral microbes rather than single species from a single
genera {[}16,17{]}. Based on this discrepancy, the second question we
can answer with this meta-analysis is if there is one dominant
CRC-associated genera that can be identified across studies.

@@

In comparison to the previous meta-analysis, this study significantly
increases the total stool samples investigated, examines differences
between stool and tissue microbiota in the context of CRC, and takes a
more community centric approach rather than a biomarker focused approach
to investigating commonalities across study for the microbiota and CRC.
Importantly, this community centric approach could provide valuable
insights into the importance of accounting for the community in CRC
disease not previously provided by earlier meta-analysis studies
{[}11{]}.

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether diversity changes throughout
disease (control to adenoma to carcinoma) and if it results in an
increased relative risk (RR) for adenoma or carcinoma. We then assessed
how common CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the RR of adenoma or carcinoma. Next, using Random Forest models,
we analyzed whether the full community or only the CRC-associated genera
resulted in better model classification based on the area under the
curve (AUC). Our results suggest that the community changes as disease
severity worsens and that this community is important for disease
classification. However, since the changes in community were subtle for
adenoma we also examined what effect and sample size the studies that
were used were adequately powered for. Although we analyzed data sets
that sampled large numbers of individuals, our results indicate the
individual studies were underpowered for detecting effect size
differences of 10\% or below between the case and control groups.

@@

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} By power transforming and Z-score normalizing the
\(\alpha\)-diversity metrics for the entire data set we assessed whether
there were any broad scale community differences that could be detected
as disease severity worsened. Using linear mixed-effect models to
control for study, re-sampling of the same individual (tissue studies
only), and variable region, there was a significant decrease from
control to adenoma to carcinoma for both evenness and Shannon diversity
in stool (P-value = 0.025 and 0.043, respectively) and no
\(\alpha\)-diversity correlations in tissue (P-value \textgreater{}
0.05) {[}Figure 1{]}. We next tested whether these detectable
differences in community resulted in significant increases in RR. For
fecal samples, a decrease in evenness resulted in a significantly
increased RR for carcinoma (RR = 1.36 (1.15 - 1.61), P-value = 0.00034)
and adenoma (RR = 1.16 (1.01 - 1.34), P-value = 0.032) while a decrease
in Shannon diversity only increased the RR for carcinoma (RR = 1.33
(1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}. Interestingly, for both
adenoma and carcinoma there was no increase in RR within tissue samples
for any alpha diversity metric investigated {[}Table S1-S3{]}.

Using the Bray-Curtis distance metric, there was a significant
difference across study in the bacterial community of fecal samples
between carcinoma and controls, but not adenoma and controls {[}Table S4
\& S5{]}. For studies with unmatched tissue samples a similar trend was
observed {[}Table S3 \& S4{]} while studies with tissue samples from the
same individual (matched) had no differences {[}Table S6 \& S7{]}.

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
metrics, we assessed if the presence of specific genera resulted in a
higher RR for both stool and tissue. To investigate this we analyzed the
classically associated CRC genera, \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}
for an increase in RR. The majority of CRC-associated genera for both
feces and tissue had a significantly increased RR for carcinoma but not
for adenoma {[}Figure 3{]}. The RR effect size was greater for stool (RR
range = 1.62 - 2.37) than for tissue (RR range = 1.21 - 1.81). This
decrease may be explained by the fact that the tissue analysis included
matched samples from the same individual. In fecal samples, the RR for
carcinoma due to the presence of CRC-associated genera was greater than
either the RR associated with evenness or Shannon diversity {[}Figure 2
\& 3{]}. Additionally, the RR of carcinoma continuously increased as
individuals tested positive for more CRC-associated genera {[}Figure 3B
\& 3D{]}.

There were two significant measures for increased RR of adenoma when
investigating CRC-associated genera in stool: 1) Having a higher then
median value of \emph{Porphyromonas} (P-value = 0.023) and 2) whether
samples were positive for three CRC-associated genera (P-value = 0.022)
{[}Figure 3A{]}. With tissue, there were three significant measures for
an increased RR of adenoma: 1) being positive for one CRC-associated
genera (P-value = 0.032), 2) being positive for two CRC-associated
genera (P-value = 0.008), and 3) being positive for four CRC associated
genera (P-value = 0.039) {[}Figure 3C{]}.

@@

Our first approach using genus based models showed an AUC decrease in
model classification on the training set for both stool and tissue
studies {[}Figure S2-S3{]}. With respect to the test sets, comprised of
genera data from other studies, both the all genera model and
CRC-associated models had a similar ability to detect adenomas or
carcinomas {[}Figure S4-S6{]}. Two interesting but separate general
observations from these models were that: 1) classification of adenomas
was lower than carcinomas for both tissue and stool and 2) AUC for the
classification of carcinoma was consistently lower for the tissue models
than the stool models {[}Figure S4-S6{]}.

@@

In stool the most common genera in the genus based models belonged
mostly to resident genera such as \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} {[}Figure 6A \& B{]}. With
respect to the CRC-associated genera, \emph{Fusobacterium} was the only
genus present in adenoma while all four were present in carcinoma
{[}Figure 6A \& B{]}. Conversely, none of these CRC-associated genera
were present in the majority of studies. When we move on to the OTU
based models, the adenoma OTU models had OTUs that classified as
\emph{Ruminococcaceae} or \emph{Roseburia} present in the top 10 OTUs
for the vast majority of studies {[}Figure 6C{]}. \emph{Ruminococcaceae}
was also present in the top 10 in some studies for the carcinoma OTU
models, but it was \emph{Bacteroides} that was present in the
overwhelming majority of the carcinoma OTU stool models {[}Figure 6D{]}.

@@

Our study identifies clear differences in diversity both at the
community level and within individual genera that are present in
individuals with CRC versus those without the disease {[}Figure 1-3{]}.
Although there was a step-wise decrease in diversity from control to
adenoma to carcinoma, this did not translate into large effect sizes for
the RR of either of these two. Even though CRC-associated genera
increase the RR of carcinoma, they do not consistently increase the
relative risk of adenoma. This information suggests that these specific
genera are strongly associated with carcinoma but may not be associated
with adenoma. Although CRC-associated genera are associated with
carcinomas, our data show that by using the whole community,
classification models perform better than when only the CRC-associated
genera are included. CRC-associated genera are clearly important to
carcinoma pathogenesis but accounting for the community in which these
microbes exist can drastically increase the ability of models to make
predictions.

The data presented herein supports the driver-passenger model of the
microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
carcinoma but not necessarily adenoma. The central idea of the model is
that a single bacterium initiates an environment in which other
non-resident microbes may then be able to colonize, creating a vicious
cycle that is conducive for CRC. Both the drastically increased
carcinoma RR of CRC-associated genera versus \(\alpha\)-diversity
metrics and increasing RR with more CRC-associated genera positivity,
are highly supportive of this model. However, the initial establishment
of the driver within the system appears to be dependent on the current
community. This is supported by our finding that when adding the
community context to our models in addition to the CRC-associated
genera, the model AUC increases {[}Figure 4 \& 5{]}. Conversely, using
the present data, it is less likely that adenoma development fits this
model. The changes that occur at this timepoint are small and possibly
focal to the adenoma itself. The step-wise decrease in diversity
suggests that the adenoma community is not normal but has changed subtly
{[}Figure 1{]}. Although there appears to be localized changes that do
depend on the driver-passenger model, as supported by an increased RR
for one, two, and four positive CRC-associated genera in tissue
{[}Figure 3C{]}, there may be other processes at play that ultimately
exacerbate the condition from a subtle localized change, to a change in
the global community. The poor performance of the Random Forest models
for classifying adenoma, based only on the microbiota, would suggest
that this is the case. One potential hypothesis from these observations
is that at early stages of the disease, how the host interacts with
these subtle changes is what ultimately leads to a thoroughly
dysfunctional community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be disciminative across the different studies for
both adenoma and carcinoma models {[}Figure 6{]}. This data would
suggest that whether the non-resident bacterium is \emph{Fusobacteria}
or \emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. These observations would
also suggest that initial changes in the resident community,
specifically to \emph{Bacteroides}, \emph{Ruminococcaceae},
\emph{Ruminococcus}, and \emph{Roseburia}, carry on from adenoma to
carcinoma. Based on these observations, it is possible to hypothesize
that the early changes in the community may give rise to initial polyp
formation via interactions with the host and not necessarily via
interactions with CRC-associated genera. These changes then create new
niches in which any one of the CRC-associated genera could gain a
foothold, exacerbating the initial changes in community and facilitating
the transition from adenoma to carcinoma via a driver-passenger type
mechanism.

@@

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, \emph{et
al.} and Vogtmann, \emph{et al.} {[}29,30{]}. Additional studies not
mentioned in the reviews were obtained based on the authors' knowledge
of the literature. Studies that used tissue or feces as their sample
source for 454 or Illumina 16S rRNA gene sequencing analysis were
included. Only data sets that had sequences available for analysis were
included. Some studies did not have publicly available sequences or did
not have metadata in which the authors were able to share and were
excluded. After these filtering steps, the following studies remained:
Ahn, \emph{et al.} {[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim,
\emph{et al.} {[}22{]}, Burns, \emph{et al.} {[}27{]}, Chen, \emph{et
al.} {[}19{]}, Dejea, \emph{et al.} {[}24{]}, Flemer, \emph{et al.}
{[}12{]}, Geng, \emph{et al.} {[}28{]}, Hale, \emph{et al.} {[}18{]},
Kostic, \emph{et al.} {[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy,
\emph{et al.} {[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et
al.} {[}23{]}, and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]}
study was not included because the 90 individuals analyzed within the
study are contained within the larger Baxter study {[}16{]}.
Additionally, after sequence processing all the case samples for the
Kostic study only had 100 or less sequences remaining and was not used.
This left a total of 14 studies for which analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}26{]} and for four studies was obtained directly from
the authors {[}12,18,23,25{]}. Each study was processed using the mothur
(v1.39.3) software program {[}32{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identified and
removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}34{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.2) software package {[}35{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
normalized using the car (v2.1.5) package {[}37{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}38{]} to correct for study,
repeat sampling of individuals (tissue only), and variable region
effects. Relative risk was analyzed using both the epiR (v0.9.87) and
metafor (v2.0.0) packages {[}39,40{]} by assessing how many with and
without disease were above and below the overall median value within the
specific study. Relative risk significance testing utilized the
chi-squared test. \(\beta\)-diversity differences utilized a Bray-Curtis
distance matrix and PERMANOVA executed with the vegan (v2.4.4) package
{[}41{]}. Random Forest models were built using both the caret (v6.0.77)
and randomForest (v4.6.12) packages {[}42,43{]}. Power analysis and
estimations were made using the pwr (v1.2.1) and statmod (v1.4.30)
packages {[}44,45{]}. All figures were created using both ggplot2
(v2.2.1) and gridExtra (v2.3) packages {[}46,47{]}.
